---
title: "Requirements"
date: "27/11/2017"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
---




# Data 

### First part of Friday:
Can be found on [github](https://github.com/ComputationalRegulatoryGenomicsICL/ZENCODE_Advanced_RNA-seq_training). Please click on the `clone/download` button to download the directory or if you're git savy clone it! 

### Second part of Friday:
Can be found [here](https://www.dropbox.com/sh/xk5dc6poh8oosoo/AACd-UOXXpY8d7mrXHGXDwtia?dl=0). Please download the whole directory. Scripts and presentations are also on [github](https://github.com/ComputationalRegulatoryGenomicsICL/ZENCODE_Advanced_RNA-seq_training)

# R
Download or please update R to the latest version 3.4.2) from [here](https://www.r-project.org/).


# RStudio 

Download and install RStudio from the following link (if you already have RStudio, update to latest version): [RStudio weblink](https://www.rstudio.com/products/RStudio/#Desktop).

# R Packages
If asked to update packages `Update all/some/none [a/s/n]` - update all packages (answer a).

## Bioconductor

- Update Bioconductor

```{r eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite()
```

- The rest of the packages

```{r eval = FALSE}
biocLite("DESeq2")
biocLite("DEXSeq")
biocLite("maSigPro")
biocLite("clusterProfiler")
biocLite("org.Dr.eg.db")
```

## CRAN

```{r eval = FALSE}
install.packages("RColorBrewer")
install.packages("gplots")
install.packages("ggplot2")
install.packages("venn")
install.packages("tidyverse")
```

